---
title: "Community Composition & Diversity from a Phyloseq Object"
author: "Jonathan Rondeau-Leclaire"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    number_sections: true
    css: mgx.tutorials.css
---

<div class="callout-note">
**About this tutorial**

Once you have a `phyloseq` object (ASV/OTU table + taxonomy + sample metadata, all linked together), three analyses come up again and again: a **composition barplot** (relative abundance by taxonomic rank), an **alpha diversity** comparison (within-sample diversity across groups), and a **beta diversity** ordination (between-sample dissimilarity, e.g. a PCoA). This tutorial builds all three, one at a time, from the same publicly available phyloseq object.

It's adapted from a real analysis script, [`5.2_Figs3_4_5_Plots_byCity.R`](https://github.com/jorondo1/urban_bioaerosols/blob/main/scripts/5.2_Figs3_4_5_Plots_byCity.R), which builds a publication-grade multi-panel figure, published in STOTEN by [Poirier S., Rondeau-Leclaire J., et al 2026](https://www.sciencedirect.com/science/article/pii/S0048969726002846). That version relies on more advanced coding, so we're going to do a simplified version of one of these plots.

</div>

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
klippy::klippy(c('r'), color = 'lightblue', tooltip_message = 'Click to copy', position = c('top', 'right'), tooltip_success = 'Done')
```

# Setup

```{r env}
library(pacman)
p_load(mgx.tools, # devtools::install_github("jorondo1/mgx.tools")
       tidyverse, phyloseq, RColorBrewer, vegan, rstatix, ggpubr, kableExtra, knitr)
```

`mgx.tools` is the same in-house package used in the [DADA2 tutorial](dada2_16S_tutorial.html); a couple of its small utilities (`samdat_as_tibble()`) show up below.

## Download a toy dataset

If you don't have your own data, you can use this one from `Poirier et al. 2026`. The file actually contains three datasets: bacteria, fungi, and plant particles, pre-processed and rarefied to even sequencing depth.

```{r load-data}
poirier_data <- read_rds('https://raw.githubusercontent.com/jorondo1/urban_bioaerosols/main/data/ps_rare.ls.rds')
ps_fung <- poirier_data$FUNG
ps_fung
```

A phyloseq object bundles three tables together (the ASV/OTU count table, the taxonomy table, and the sample metadata) and keeps them aligned. Pull out just the sample metadata as a tibble, which is all we need to see what we can group by:

```{r peek-samdat}
samdat_as_tibble(ps_fung) %>%
  select(Sample, city, time, date) %>%
  head()
```

Three sampling cities (`city`), three sampling periods (`time`: Spring/Summer/Fall), each spread across a handful of `date`s. We'll aggregate across `date`, keeping only `city` and `time`.

If you have your own data, you can adapt the code below to your own variables, depending on which metadata categories you want to show separately.

# Analyses {.tabset}

## Relative abundance

**Objective: ** plot the mean relative abundance for each season within each city, while showing only the top 12 taxa and bundling the rest in an "Others" category.

### Melt the phyloseq object

`psmelt()` turns a phyloseq object into one long-format table, one row per ASV × sample, with the taxonomy and sample metadata already joined in. 

```{r melt}
melted_ps <- psflashmelt(ps_fung) 
dim(melted_ps)
```

You can peak into it using `View(melted_ps)` on RStudio, you'll see it's quite a large data frame! Let's tidy it up for ggplot.

### Aggregate to Genus level and compute relative abundance

Pick the taxonomic rank you want to plot, e.g. `Genus`, and aggregate every ASV within each **sample** down to its `Genus`, then normalize within each sample before averaging into `city`/`time` groups:

```{r aggregate}
relab_df <- melted_ps %>%
  
  # Summarise at desired rank:
  group_by(Sample, city, time, Genus) %>%  # unspecified variables will be dropped!
  summarise(Abundance = sum(Abundance), .groups = 'drop') %>%
  
  # compute relative abundances:
  group_by(Sample) %>%
  mutate(relAb = Abundance / sum(Abundance)) %>%
  
  # compute mean relative abundance by city X time combinations for each Genus:
  group_by(city, time, Genus) %>%
  summarise(relAb = mean(relAb), .groups = 'drop') %>% 
  
  # only keep non-zero rows:
  filter(relAb>0)

relab_df
```

<div class="callout-note">
We normalize **within each sample first** (`relAb = Abundance / sum(Abundance)`, grouped by `Sample`), then average those proportions across every sample in a `city`/`time` group. This matters whenever samples don't all have the same sequencing depth: without it, a sample with 200,000 reads would dominate the group average over one with 20,000 reads, even if they're biologically equivalent.

If your phyloseq object is already **rarefied** to even depth (as this one is), every sample already carries equal weight, so you can skip the per-sample step and just sum raw abundances directly within `city`/`time` and normalize once — same result, one less `group_by()`.
</div>

### Collapse rare genera into "Other"

`relab_df` has `r n_distinct(relab_df$Genus)` distinct genera, wayyy too many for a readable legend. `forcats::fct_lump_n()` keeps the `n` highest-weighted levels and lumps everything else into a single `"Other"` category. We'll plot 12 genera.

```{r lump}
relab_df <- relab_df %>%
  mutate(Genus = fct_lump_n(Genus, n = 12, w = relAb, other_level = "Other")) %>%
  group_by(city, time, Genus) %>%
  summarise(relAb = sum(relAb), .groups = 'drop')

relab_df
```

### Plot tweaks [optional]

Two small tweaks make for a nicer plot: order `time` chronologically instead of alphabetically, and order `Genus` alphabetically too, throwing `"Unclassified"` and `"Other"` at the end of the list. You can adapt this to whatever data you have, because usually we want to present our categories in a certain order, and R doesn't know what order that is.

```{r factors}
genus_levels <- levels(relab_df$Genus)
# keep whichever of these two are actually present, in this order:
special_levels <- intersect(c("Unclassified", "Other"), genus_levels)
ordered_levels <- c(sort(setdiff(genus_levels, special_levels)), special_levels)

relab_df <- relab_df %>%
  mutate(
    time  = factor(time, levels = c("Spring", "Summer", "Fall")),
    Genus = factor(Genus, levels = ordered_levels)
  )
```

Since we might end up showing more taxa than a Brewer palette can offer (`RColorBrewer` palettes top out at 12 colours), we expand one by interpolating between its colours. 

```{r palette}
taxa_pal <- colorRampPalette(brewer.pal(9, "Set1"))(nlevels(relab_df$Genus))
taxa_pal
```

You can also build your own palette, obviously! Or even modify that one.

```{r}
taxa_pal[13] <- "grey80" # (12+1)th position is "Others"
taxa_pal[12] <- "grey50" # 12th position is Unclassified
```


### Plot !

We'll put `time` on the x-axis, and facet our plot to have one panel per `city`. You can do other designs, lookup the doc for `facet_wrap()` and `facet_grid()`.

```{r plot-relab, fig.width=9, fig.height=4.5}
ggplot(relab_df, aes(x = time, y = relAb, fill = Genus)) +
  geom_col(position = "stack") +
  facet_grid(~city) +
  scale_fill_manual(values = taxa_pal) +
  labs(x = NULL, y = "Mean relative abundance", fill = "Genus") +
  theme_light()
```

<details>
<summary>
What if I want to show more categories?
</summary>
If you had a third variable (let's call it `condition`), one way to modify that plot would be to add it to the faceting variables, e.g. `facet_grid(condition~city)`. You would still get one column-faceting by city, but also one row-faceting by condition now; the y axis would appear as many times as you have conditions.
</details>

Notice the empty Quebec-Summer panel: there's no data for that combination. By default, `facet_grid()` will show it, empty. If you wanted to hide it, you could add the parameted `scales = 'free_x'` so each panel will only contain the x values for which it has data; it really depends on your goal. 

<div class="callout-danger">
**Sanity check!**

Imagine you needed add a third grouping variable to the data code above (say, `date`). But you forgot to add that variable to the plot! `relab_df` would then have *more than one row* per `city`/`time`/`Genus`, which would add up to more than one. To double-check the data is fine,

```{r sanity-check}
# 1. Exactly one row per (city, time, Genus)? This should return 0 rows.
relab_df %>% count(city, time, Genus) %>% filter(n > 1)

# 2. Does every bar actually sum to 1?
relab_df %>% group_by(city, time) %>% summarise(total = sum(relAb), .groups = 'drop')
```

To make sure you don't miss it, use  `position = "stack"` (the default for `geom_col()`, spelled out here on purpose) which **sums** every row sharing an `x`/`fill` pair rather than hiding them.

</div>


## Alpha diversity

Alpha diversity is diversity *within* a sample: how many taxa there are, and how evenly they're distributed. We'll use the Shannon index, a very commonly reported index, and compare it across seasons, within each city.

### Rarefy, then compute Shannon diversity

According to the latest research ([Schloss, 2024](https://journals.asm.org/doi/10.1128/msphere.00354-23)), diversity indices needs to be rarefied to the same sequencing depth to be comparable. Otherwise, a deeper sample could look "more diverse" simply because it was sequenced more. 

```{r rarefy-alpha}
set.seed(1) # for reproducibility
ps_fung_rare <- rarefy_even_depth(ps_fung, rngseed = 1)
```

<div class="callout-note">
This toy dataset is already rarefied, so the step above changes nothing really. Yours might not be! 

Actually, ideally we'd also rarefy *many* times and average each index across iterations, since any single random draw can be unlucky; `mgx.tools::rarefy_diversity()` does exactly that (and much more), but requires a lot of processing time.
</div>

`mgx.tools::estimate_diversity()` computes the index you want directly from a phyloseq object, generating one value per sample.

```{r shannon}
shannon_df <- mgx.tools::estimate_diversity(ps_fung_rare, "Shannon") %>%
  enframe(name = "Sample", value = "Shannon")
```

To create a plot-friendly data frame, we'll extract our `sample_data` from the phyloseq object and add this Shannon column to it.
```{r }
alpha_df <- mgx.tools::samdat_as_tibble(ps_fung_rare) %>%
  select(Sample, city, time) %>%
  left_join(shannon_df, by = "Sample") %>%
  mutate(time = factor(time, levels = c("Spring", "Summer", "Fall")))

alpha_df
```

<div class="callout-note">
For alternative indices, `estimate_diversity()` currently supports Richness, Shannon, Simpson or Tail. If you are interested in Hill numbers, Hill order 1 is `exp(Shannon)` and Hill order 2 is `1/Simpson`; you can create them using `mutate()`.
</div>

### Plot

Let's make a violin plot !

```{r plot-alpha, fig.width=8, fig.height=4}
alpha_plot <- alpha_df %>% 
  ggplot(aes(x = time, y = Shannon, fill = time)) +
  geom_violin(linewidth = 0.2, draw_quantiles = 0.5) +
  facet_grid(~city) +
  labs(x = NULL, y = "Shannon index", fill = "Season") +
  theme_light()
alpha_plot 
```

### Wilcoxon tests, corrected for multiple comparisons

We use `rstatix::wilcox_test()` to run one test per pair of seasons, per group (here, `city`). `adjust_pvalue()` then corrects those p-values (Benjamini-Hochberg, i.e. false discovery rate):

```{r alpha-stats}
alpha_stats <- alpha_df %>%
  group_by(city) %>%
  wilcox_test(Shannon ~ time) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

alpha_stats
```

<div class='callout-danger'>
Notice that these only test **within city**, not between cities! This is what `group_by()` serves.
</div>

Package it as an html table:

```{r alpha-kable}
alpha_kable <- alpha_stats %>%
  select(city, group1, group2, p, p.adj, p.adj.signif) %>%
  kable(caption = "Wilcoxon rank-sum test: Shannon diversity by season, within each city (BH-adjusted)") %>%
  kable_styling(full_width = FALSE)

alpha_kable
```

Or add it to the plot! `rstatix::add_xy_position()` reuses `alpha_stats` directly to work out where each bracket should sit (and which two `time` levels it spans), and `ggpubr::stat_pvalue_manual()` draws it — `city` being a column of `alpha_stats` is what lets it land brackets in the right facet panel. `hide.ns = TRUE` skips drawing a bracket for the non-significant comparisons.

```{r alpha-plot-with-stats, fig.width=9, fig.height=4.5}
alpha_stats_pos <- alpha_stats %>% add_xy_position(x = "time")

alpha_plot <- alpha_plot +
  stat_pvalue_manual(alpha_stats_pos, label = "p.adj.signif", hide.ns = TRUE)
```

### Save your output!

```{r alpha-save, eval = FALSE}
save_kable(alpha_kable, "alpha_wilcox.html")
ggsave(alpha_plot, "alpha_div.pdf", bg = 'white', 
       width = 2200, height = 2000, units = 'px', dpi = 220)
```

`save_kable()` writes the table out as a standalone `.html` file you can share or embed elsewhere. It's also easy to copy from, to paste into Word Office for example.

## Beta diversity

Beta diversity compares dissimilarity *between* samples. We'll compute Bray-Curtis distance on variance-stabilized counts, visualize it with a PCoA, and test it with a PERMANOVA.

### Rarefy, variance-stabilize, and compute Bray-Curtis

```{r rarefy-beta}
set.seed(1)
ps_fung_rare <- rarefy_even_depth(ps_fung, rngseed = 1)
```

Same caveat as for alpha diversity: one rarefaction draw here for simplicity, `mgx.tools::rarefy_diversity()` for the iterate-and-average version.

`mgx.tools::compute_pcoa()` does three things in one call: 
1. variance-stabilizes the counts (using DESeq2's VST: handles the mean-variance relationship in count data better than a plain log-transform);
2. computes a Bray-Curtis distance matrix on the transformed counts
3. then runs classical multidimensional scaling (PCoA) to get 2D coordinates:

```{r pcoa}
bray_pcoa <- compute_pcoa(ps_fung_rare, dist = "bray", vst = TRUE)
names(bray_pcoa)
```

It returns a list: `metadata` (sample data with two `PCo1`/`PCo2` columns added, for plotting), `eig` (eigenvalues, for the % variance explained by each axis, which we'll add to the plot), and `dist.mx` (the distance matrix itself — we'll need it for the PERMANOVA below).

### Plot the PCoA

Prep the data:

```{r}
# Dataset levels: 
pcoa_df <- bray_pcoa$metadata %>%
  mutate(time = factor(time, levels = c("Spring", "Summer", "Fall")))

# One decimal is enough for the PCo contribution %:
eig_pct <- round(100 * bray_pcoa$eig[1:2] / sum(bray_pcoa$eig), 1)

```

Plot ! 

```{r plot-beta, fig.width=6.5, fig.height=4.5}
beta_plot <- pcoa_df %>% 
  ggplot(aes(x = PCo1, y = PCo2, colour = time, fill = time)) +
  geom_point(size = 2, shape = 21, colour = 'black') +
  stat_ellipse(geom = "polygon", alpha = 0.15, level = 0.95) +
  
  # Add the axis PCo % contribution 
  labs(x = paste0("PCo1 (", eig_pct[1], "%)"),
       y = paste0("PCo2 (", eig_pct[2], "%)"),
       colour = "Season", fill = "Season") +
  theme_light()
beta_plot
```

Samples separate cleanly by season here, We have a strong beta diversity signal. PERMANOVA below tests this formally. You could just as easily colour or facet by `city` instead.

<div class='callout-danger'>
NOTE: if you wanted to show one PCoA per city, don't simply split `bray_pcoa$metadata` in three. You should be doing one independent PCoA per city. 
</div>

### PERMANOVA

`vegan::adonis2()` tests whether group membership explains a significant share of the distance matrix. We model both `city` and `time` (season), see which variable explains the most variance in beta diversity.

```{r permanova}
set.seed(1)
permanova <- adonis2(
  bray_pcoa$dist.mx ~ city + time,  # the linear model
  data = bray_pcoa$metadata,        # the dataset
  permutations = 999,               # the precision of p-values
  by = "margin")                     # how the model should be handled (read on that!)
permanova
```

You can bump up the permutations to 9999 for a better p-value estimate.

```{r beta-kable}
beta_kable <- permanova %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  filter(term != "Total") %>%
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "PERMANOVA: Bray-Curtis ~ city + time") %>%
  kable_styling(full_width = FALSE)

beta_kable
```

### Save your output!

```{r beta-save, eval = FALSE}
save_kable(beta_kable, "beta_permanova.html")
ggsave(pcoa_df, "beta_pcoa.pdf", bg = 'white', 
       width = 2200, height = 2000, units = 'px', dpi = 220)
```

Both `city` and `time` explain a significant share of the variance (the `R2` column), consistent with the clear seasonal clustering in the plot above.

## {-}
