---
title: "Community Composition & Diversity from a Phyloseq Object"
author: "Jonathan Rondeau-Leclaire"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    number_sections: true
    css: mgx.tutorials.css
---

<div class="callout-note">
**About this tutorial**

Once you have a `phyloseq` object (ASV/OTU table + taxonomy + sample metadata, all linked together), three analyses come up again and again: a **composition barplot** (relative abundance by taxonomic rank), an **alpha diversity** comparison (within-sample diversity across groups), and a **beta diversity** ordination (between-sample dissimilarity, e.g. a PCoA). This tutorial builds all three, one at a time, from the same publicly available phyloseq object.

It's adapted from a real analysis script, [`5.2_Figs3_4_5_Plots_byCity.R`](https://github.com/jorondo1/urban_bioaerosols/blob/main/scripts/5.2_Figs3_4_5_Plots_byCity.R), which builds a publication-grade multi-panel figure, published in STOTEN by [Poirier S., Rondeau-Leclaire J., et al 2026](https://www.sciencedirect.com/science/article/pii/S0048969726002846). That version relies on more advanced coding, so we're going to do a simplified version of one of these plots.

</div>

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
klippy::klippy(c('r'), color = 'lightblue', tooltip_message = 'Click to copy', position = c('top', 'right'), tooltip_success = 'Done')
```

# Setup

```{r env}
library(pacman)
p_load(mgx.tools, # devtools::install_github("jorondo1/mgx.tools")
       tidyverse, phyloseq, RColorBrewer)
```

`mgx.tools` is the same in-house package used in the [DADA2 tutorial](dada2_16S_tutorial.html); a couple of its small utilities (`samdat_as_tibble()`) show up below.

## Download a toy dataset

If you don't have your own data, you can use this one from `Poirier et al. 2026`. The file actually contains three datasets: bacteria, fungi, and plant particles, pre-processed and rarefied to even sequencing depth.

To download, make sure you are in your working directory (or that you are working using RStudio Projects), otherwise the `data/` directory will get created somewhere (!) on your computer.

```{r load-data}
# create the directory
dir.create('data', showWarnings = FALSE)
ps_path <- 'data/ps_rare.ls.rds' # <- you can change this depending on your directory structure

# Download the .rds object
if (!file.exists(ps_path)) {
  download.file(
    'https://raw.githubusercontent.com/jorondo1/urban_bioaerosols/main/data/ps_rare.ls.rds',
    ps_path, mode = 'wb'
  )
}

# Pull the fungal phyloseq database only
ps_fung <- read_rds(ps_path)$FUNG
ps_fung
```

A phyloseq object bundles three tables together (the ASV/OTU count table, the taxonomy table, and the sample metadata) and keeps them aligned. Pull out just the sample metadata as a tibble, which is all we need to see what we can group by:

```{r peek-samdat}
samdat_as_tibble(ps_fung) %>%
  select(Sample, city, time, date) %>%
  head()
```

Three sampling cities (`city`), three sampling periods (`time`: Spring/Summer/Fall), each spread across a handful of `date`s. We'll aggregate across `date`, keeping only `city` and `time`.

If you have your own data, you can adapt the code below to your own variables, depending on which metadata categories you want to show separately.

# Analyses {.tabset}

## Relative abundance

### Melt the phyloseq object

`psmelt()` turns a phyloseq object into one long-format table, one row per ASV × sample, with the taxonomy and sample metadata already joined in. 

```{r melt}
melted_ps <- psmelt(ps_fung) 
dim(melted_ps)
```

You can peak into it using `View(melted_ps)` on RStudio, you'll see it's quite a large data frame! Let's tidy it up for ggplot.

### Aggregate to Genus level and compute relative abundance

Pick the taxonomic rank you want to plot, e.g. `Genus`, and aggregate every ASV within each **sample** down to its `Genus`, then normalize within each sample before averaging into `city`/`time` groups:

```{r aggregate}
relab_df <- melted_ps %>%
  # Summarise at desired rank:
  group_by(Sample, city, time, Genus) %>%  # unspecified variables will be dropped!
  summarise(Abundance = sum(Abundance), .groups = 'drop') %>%
  # compute relative abundances:
  group_by(Sample) %>%
  mutate(relAb = Abundance / sum(Abundance)) %>%
  # compute mean relative abundance by city X time combinations for each Genus:
  group_by(city, time, Genus) %>%
  summarise(relAb = mean(relAb), .groups = 'drop') %>% 
   # only keep non-zero rows:
  filter(relAb>0)

relab_df
```

<div class="callout-note">
We normalize **within each sample first** (`relAb = Abundance / sum(Abundance)`, grouped by `Sample`), then average those proportions across every sample in a `city`/`time` group. This matters whenever samples don't all have the same sequencing depth: without it, a sample with 200,000 reads would dominate the group average over one with 20,000 reads, even if they're biologically equivalent.

If your phyloseq object is already **rarefied** to even depth (as this one is), every sample already carries equal weight, so you can skip the per-sample step and just sum raw abundances directly within `city`/`time` and normalize once — same result, one less `group_by()`.
</div>

### Collapse rare genera into "Other"

`relab_df` has `r n_distinct(relab_df$Genus)` distinct genera, wayyy too many for a readable legend. `forcats::fct_lump_n()` keeps the `n` highest-weighted levels and lumps everything else into a single `"Other"` category. We'll plot 12 genera.

```{r lump}
relab_df <- relab_df %>%
  mutate(Genus = fct_lump_n(Genus, n = 12, w = relAb, other_level = "Other")) %>%
  group_by(city, time, Genus) %>%
  summarise(relAb = sum(relAb), .groups = 'drop')

relab_df
```

### Plot tweaks [optional]

Two small tweaks make for a nicer plot: order `time` chronologically instead of alphabetically, and order `Genus` alphabetically too, throwing `"Unclassified"` and `"Other"` at the end of the list. You can adapt this to whatever data you have, because usually we want to present our categories in a certain order, and R doesn't know what order that is.

```{r factors}
genus_levels <- levels(relab_df$Genus)
# keep whichever of these two are actually present, in this order:
special_levels <- intersect(c("Unclassified", "Other"), genus_levels)
ordered_levels <- c(sort(setdiff(genus_levels, special_levels)), special_levels)

relab_df <- relab_df %>%
  mutate(
    time  = factor(time, levels = c("Spring", "Summer", "Fall")),
    Genus = factor(Genus, levels = ordered_levels)
  )
```

Since we might end up showing more taxa than a Brewer palette can offer (`RColorBrewer` palettes top out at 12 colours), we expand one by interpolating between its colours. 

```{r palette}
taxa_pal <- colorRampPalette(brewer.pal(9, "Set1"))(nlevels(relab_df$Genus))
taxa_pal
```

You can also build your own palette, obviously! Or even modify that one.

### Plot !

Our goal is to show one column per `city`/`time` combination. We'll put `time` on the x-axis, and facet our plot to have one panel per `city`. You can do other designs, lookup the doc for `facet_wrap()` and `facet_grid()`.
Imagine you later add a third grouping variable to the data code above (say, `date`) but forget to also reflect it in the plot. `relab_df` would then have *more than one row* per `city`/`time`/`Genus`, which would add up to more than one. To make sure you don't miss it, use  `position = "stack"` (the default for `geom_col()`, spelled out here on purpose) which **sums** every row sharing an `x`/`fill` pair rather than hiding them.

```{r plot-relab, fig.width=9, fig.height=4.5}
ggplot(relab_df, aes(x = time, y = relAb, fill = Genus)) +
  geom_col(position = "stack") +
  facet_grid(~city, ) +
  scale_fill_manual(values = taxa_pal) +
  labs(x = NULL, y = "Mean relative abundance", fill = "Genus") +
  theme_light()
```

Again, you can change this and even add a third variable in the facet if you want.

Notice the empty `Quebec`/`Summer` panel — there's simply no data for that combination in this dataset. By default, `facet_grid()` will show it empty. If you wanted to hide it, you could add the parameted `scales = 'free_x'` so each panel will only contain the x values for which it has data; it really depends on your goal. 

## Alpha diversity

*Coming in a future update to this tutorial.*

Alpha diversity (e.g. Shannon index) compares diversity *within* each sample, then contrasts it across groups — typically with a boxplot or violin plot per `time`, faceted or colored by `city`. In the source project, the values themselves come from a separate script, [`3.1_diversityMetrics.R`](https://github.com/jorondo1/urban_bioaerosols/blob/main/scripts/3.1_diversityMetrics.R), which calls `mgx.tools::estimate_diversity()` on each phyloseq object to build a `Sample`-level diversity table before any plotting happens.

## Beta diversity

*Coming in a future update to this tutorial.*

Beta diversity compares dissimilarity *between* samples (e.g. Bray-Curtis distance) and is usually visualized as an ordination — a PCoA scatterplot colored by group, often with confidence ellipses. In the source project this is also built in [`3.1_diversityMetrics.R`](https://github.com/jorondo1/urban_bioaerosols/blob/main/scripts/3.1_diversityMetrics.R), via `mgx.tools::compute_pcoa()`.

## {-}

---

*This tutorial fulfills the "relative abundance matrix analyses" item from the [setup guide's roadmap](https://jorondo1.github.io/mgx.tutorials/init_ip34.html#coming-soon); the alpha/beta diversity tabs above will fill in as that work continues.*
